---
title: "Introduction to dcmodify"
author: "Mark van der Loo and Edwin de Jonge"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to dcmodify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### A first statement

In the iris dataset, replace `Sepal.Width` with 4 value if it exceeds 4.
```{r,eval=FALSE}
library(dcmodify)
library(magrittr)
iris %<>% modify_so( if(Sepal.Width > 4 ) Sepal.Width <- 4 )
```

### Why this package

Data cleaning work flows or scripts typically contain a lot of 'if this do that'
type of statements. Such statements are typically condensed expert knowledge.
With this package, such 'data modifying rules' are taken out of the code and
become instead parameters to the work flow. This allows you to maintain, document
and reason about data modification rules separately from the flow of your programme.

This means you, the expert, can focus on the content and let R do the work.


### Basic workflow

The workflow of `dcmodify` is designed to take two concerns of your hands. The first concern is how to implement the many ideas and rules that define how and when to modify data. The second concern is related to how to apply such rules to your data. We therefore introduce two nouns and one verb that govern the basic workflow.

- data: This is your data, currently this must be stored in a `data.frame`.
- `modifier`: This is an object that stores (conditional) data modification rules.
- `modify`: This is a function that applies the rules in a modifier to your data.

Here's an example using the `retailers` data set from the [validate](https://cran.r-project.org/package=validate) package. 
```{r}
data("retailers", package="validate")
head(retailers[-(1:2)],3)
```

First we define a set of modifying rules, using `modifier`.
```{r}
library(dcmodify)
m <- modifier(
  if (other.rev < 0) other.rev <- -1 * other.rev
  , if ( is.na(staff.costs) ) staff.costs <- mean(staff.costs)
)
```
Next, the rules can be applied to our data.
```{r}
ret1 <- modify(retailers,m)
```

Alternatively, if you're a fan of the [magrittr](https://cran.r-project.org/package=magrittr), package you can do this
```{r,eval=FALSE}
library(magrittr)
ret2 <- retailers %>% modifier(m)
```
or even
```{r,eval=FALSE}
retailers %<>% modify_so(
  if ( other.rev < 0) other.rev <- -1 * other.rev
  , if ( is.na(staff.costs) ) staff.costs <- mean(staff.costs)
)
```
Here, the `%<>%` operator makes sure that the original dataset gets overwritten, and `modify_so` is a shortcut function for defining modificaton rules in-line.

### Handling missing values

The rules you define in a `modifier` are executed on records where the conditions yields `TRUE`. In R this poses the problem on what to do when in a record the condition evaluates to `NA`. For example, the condition
```
other.rev < 0
```
in the first rule of `m` above evaluates to `NA` in the first record of the `retailers` dataset. Such cases are handled by treating it as if the condition evaluated to `FALSE`.


### Exporting and importing rules from file

Modifier rules can also be defined and stored outside of the R script through the use of YAML files. Defining a YAML file can be done by hand, or by exporting an existing modifier object via `export_yaml` or `as_yaml`. Exporting the modifier defined in the [Basic workflow] section would look as follows:
```{r,eval=FALSE}
export_yaml(m, "myrules.yaml")
```
This code will create a YAML file with the following content: 
```
rules:
- expr: if (other.rev < 0) other.rev <- -1 * other.rev
  name: M1
  label: ''
  description: ''
  created: 2021-07-29 16:57:00
  origin: command-line
  meta: []
- expr: if (is.na(staff.costs)) staff.costs <- mean(staff.costs)
  name: M2
  label: ''
  description: ''
  created: 2021-07-29 16:57:00
  origin: command-line
  meta: []
```
Out of all these keys only `rules:` and `expr:` are required, all others are optional. 

Once a YAML file is created, `modifier` can read the modification rules from the file and store it as a modifier object. For this the `.file` argument is used:
```{r,eval=FALSE}
m <- modifier(.file = "myrules.yaml")
```
Using separate files for the storage of rules has the advantage that the same set of rules can be easily shared across many different scripts.

### Exporting and importing rules as data frame

Modifier rules can also be defined and stored inside the R script through the use of data frames. Defining a data frame can be done by hand, or by exporting an existing modifier object via `as.data.frame`. Exporting the modifier defined in the [Basic workflow] section would look as follows:
```{r,eval=FALSE}
df <- as.data.frame(m)
```
This code will create a data frame with the following columns: 
```
rule: "if (other.rev < 0) other.rev <- -1 * other.rev" "if (is.na(staff.costs)) staff.costs <- mean(staff.costs)"
name: "M1" "M2"
label:"" "" 
description:"" "" 
created: 2021-07-29 16:57:00 2021-07-29 16:57:00
origin: "command-line" "command-line"
meta: [] []
```
Out of all these columns only `rule` `name` and `description` are required, all others are optional. 

Once a data frame is created, `modifier` can read the modification rules from the data frame and store it as a modifier object. For this the `.data` argument is used:
```{r,eval=FALSE}
m <- modifier(.data = df)
```


### Options

By default, most options are taken from `validate` options (see `validate::.PKGOPT()`).
Options can be set by passing them as arguments to `modify`. For example, by setting `sequential` to `FALSE`,
you specify that assignments should be independent from previous assignments:
```{r,eval=FALSE}
df <- data.frame(
  a = 1:2,
  b = 3:4
)

m <- modifier(if (a == 1 & b == 3) { a <- 10; b <- 30 })

# Only a is modified.
modify(df, m)

# Both a and b are modified.
modify(df, m, sequential = FALSE)
```

Available options include:

* `sequential`
* `na.condition`


### Performance, and a glimpse under the hood.

You, the user can assume that the rules are evaluated record-by-record. In
reality, the package is smart enough to analyse the rules a little bit and to
make sure they can be evaluated in a vectorized manner. This way explicit (and slow)
R-loops are avoided as much as possible.

In short, when you call `modify`, or `modify_so`, the following steps are performed.

1. The rules are transformed to statements that can be executed in a vectorized manner by R.
2. If any macros present, they are inserted into the statements.
3. For each assignment, the conditions under which they should be executed are collected.
4. The conditions are evaluated and assignments are executed on a selection of the data.


### Difference with dplyr::mutate

The functionality of this package resembles `dplyr::mutate`, since it also 
allows one to specify data mutations on data frames (or other tabular data 
objects). The dplyr package is especially useful for interactive use and also for use in programming through 'underscored' functions such as `mutate_`.

The `dcmodify` package has been developed with a production street in 
mind where similar data sets are processed frequently. By taking the modifying 
rules out of the software, R programmers can build an application that allows 
users that are less knowledgeable about programming to specify their modification
rules. 

### Logging changes

It can be interesting to study the effect of a certain set of data modifying 
rules. The [lumberjack package](https://CRAN.R-project.org/package=lumberjack) is
capable of tracking changes in data.

To start logging data you need to replace the magrittr pipe (`%>%`) with the 
lumberjack operator `%>>%` and insert some logging commands into the pipeline.
```{r,eval=TRUE}
library(lumberjack)
# add primary key so cellwise changes can be traced
women$id <- letters[1:15]

out <- women %>>%
  start_log( cellwise$new(key="id") ) %>>%
  modify_so( if (height < mean(height)) height <- mean(height) ) %>>%
  dump_log()

# The log is written to file.
read.csv("cellwise.csv") %>>% head()
```


### Current limitations

Conditional statements including `else` are not supported yet. Rules containing
`if() else` are ignored with a warning.
